---
title: "Survival Analysis and Machine Learning Methodologies for Bone Marrow Cancer"
output:
  html_document:
    toc: yes
    theme: united
  pdf_document:
    toc: yes
author: Yitian Cai, Rudra Guin, Jorge Portugal
date: "12/17/2021"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Introduction

For this project, our group chose to perform statistical analysis using the Clinical Data Set from the Clinical Package of bone marrow cancer data. It was collected from the National Cancer Institute (NCI) genomic data portal. The version of the dataset we used was released by October 29, 2021. The bone marrow cancer can damage the bones, immune system, kidneys, and red blood cell count. The dataset contains 16,029 different cases of bone marrow cancer. We planned on using it for survival analysis and machine learning methodologies.

# Data Description
The dataset contains 47 clinical variables. At the beginning, we tried to include the biospecimen data in order to increase the dimension of our dataset, but we found that most of those genomics data were missing, which would not provide us too much help. So, we decided to stick with the clinical data of bone marrow. Vital status means whether the patients were alive or dead at the time the data were collected. Our dataset contains 9,036 alive cases and 3,551 dead cases. Others were either missing or recorded as "unknown" or "not reported".

# Data Preprocessing

The initial dataset contains several missing values among the different features. So, before going into analysis and modeling, we first clean up and pre-process the data. First, we remove all the observations that have missing values for vital status, because vital status is the one we wish to forecast or predict. Then, we decided to remove variables with more than 8,000 missing values. Because it means that more than half of the observations in these variables were missing, which we believe will cause too much inaccuracy after data imputation. After that, we used mice() to impute those missing values. The mice package implements a method to deal with missing data and it generates Multivariate Imputations by Chained Equations. And finally, we randomly split the dataset into two where the training set contains 70% of the data, and the test set contains the rest 30% of the data.

```{r, echo=FALSE, message=FALSE}
# Necessary libraries
library(mice)
library(fastDummies)
library(neuralnet)
library(caret)
library(randomForest)
library(caTools)
set.seed(625)
```


```{r, echo=FALSE}
# Loading the clinical data
data <- read.delim(file = 'clinical_bone.tsv', sep = '\t', header = TRUE)
data[data == "\'--"] <- "NULL"
is.na(data) <- data == "NULL"
data <- data[, colSums(is.na(data)) < nrow(data)]
#names(data)

# Transforming unknown values into NAs
data[data=="not reported"] <- NA
data[data=="Not Reported"] <- NA
data[data=="unknown"] <- NA
data[data=="Unknown"] <- NA
data[data=="not allowed to collect"] <- NA
#summary(data)
```

```{r, echo=FALSE}
# Tranforming the type of variables into numerical or categorical
data <- subset(data, select = -c(case_id, case_submitter_id))
data <- transform(
  data,
  project_id=as.factor(project_id),
  age_at_index=as.numeric(age_at_index),
  age_is_obfuscated=as.factor(age_is_obfuscated),
  cause_of_death=as.factor(cause_of_death),
  days_to_birth=as.numeric(days_to_birth),
  days_to_death=as.numeric(days_to_death),
  ethnicity=as.factor(ethnicity),
  gender=as.factor(gender),
  race=as.factor(race),
  vital_status=as.factor(vital_status),
  year_of_birth=as.numeric(year_of_birth),
  year_of_death=as.numeric(year_of_death),
  age_at_diagnosis=as.numeric(age_at_diagnosis),
  ann_arbor_b_symptoms=as.factor(ann_arbor_b_symptoms),
  ann_arbor_extranodal_involvement=as.factor(ann_arbor_extranodal_involvement),
  ann_arbor_pathologic_stage=as.factor(ann_arbor_pathologic_stage),
  burkitt_lymphoma_clinical_variant=as.factor(burkitt_lymphoma_clinical_variant),
  classification_of_tumor=as.factor(classification_of_tumor),
  days_to_diagnosis=as.numeric(days_to_diagnosis),
  days_to_last_follow_up=as.numeric(days_to_last_follow_up),
  days_to_last_known_disease_status=as.numeric(days_to_last_known_disease_status),
  inss_stage=as.numeric(inss_stage),
  iss_stage=as.factor(iss_stage),
  last_known_disease_status=as.factor(last_known_disease_status),
  method_of_diagnosis=as.factor(method_of_diagnosis),
  morphology=as.factor(morphology),
  primary_diagnosis=as.factor(primary_diagnosis),
  prior_malignancy=as.factor(prior_malignancy),
  prior_treatment=as.factor(prior_treatment),
  progression_or_recurrence=as.factor(progression_or_recurrence),
  site_of_resection_or_biopsy=as.factor(site_of_resection_or_biopsy),
  synchronous_malignancy=as.factor(synchronous_malignancy),
  tissue_or_organ_of_origin=as.factor(tissue_or_organ_of_origin),
  tumor_grade=as.factor(tumor_grade),
  year_of_diagnosis=as.numeric(year_of_diagnosis),
  days_to_treatment_end=as.numeric(days_to_treatment_end),
  days_to_treatment_start=as.numeric(days_to_treatment_start),
  initial_disease_status=as.factor(initial_disease_status),
  number_of_cycles = as.factor(number_of_cycles),
  regimen_or_line_of_therapy = as.factor(regimen_or_line_of_therapy),
  therapeutic_agents=as.factor(therapeutic_agents),
  treatment_anatomic_site=as.factor(treatment_anatomic_site),
  treatment_dose=as.factor(treatment_dose),
  treatment_intent_type=as.factor(treatment_intent_type),
  treatment_or_therapy=as.factor(treatment_or_therapy),
  treatment_outcome=as.factor(treatment_outcome),
  treatment_type=as.factor(treatment_type)
)
```



```{r, echo=FALSE}
# Remove variables with missing values in vital status
data2 <- data[!is.na(data$vital_status), ]

# Remove variables with more than 8,000 missing values
rem <- colSums(is.na(data)) > 8000
data2 <- subset(data2, select = -c(age_is_obfuscated, cause_of_death, days_to_death, year_of_birth, year_of_death, ann_arbor_b_symptoms, ann_arbor_extranodal_involvement, ann_arbor_pathologic_stage, burkitt_lymphoma_clinical_variant, classification_of_tumor, days_to_diagnosis, inss_stage, last_known_disease_status, method_of_diagnosis, prior_malignancy, prior_treatment, progression_or_recurrence, synchronous_malignancy, tumor_grade, year_of_diagnosis, days_to_treatment_end, days_to_treatment_start, initial_disease_status, number_of_cycles, regimen_or_line_of_therapy, therapeutic_agents, treatment_anatomic_site, treatment_dose, treatment_intent_type, treatment_or_therapy, treatment_outcome, treatment_type))

# summary(data)
```

```{r, echo=FALSE}
# Impute missing values
data_imp <- mice(data2, m=5, method = "sample", printFlag = FALSE)
```


# Analysis

## Survival Analysis

We first conducted some survival analysis 

```{r}

```

## Machine Learning Modeling and Analysis

We also performed some machine learning modeling and evaluated the effectiveness of the models' prediction power of the vital status of certain patients. We implemented a random forest model and a neural network model.

### Random Forest

```{r, echo=FALSE}
# Random Forest:

data_subset_rf <- complete(data_imp)
#summary(data_subset_rf)
# Remove variables that still have missing values
data_subset_rf <- subset(data_subset_rf, select = -c(morphology, primary_diagnosis))

# Apply random forest to the data
sample <- sample.split(data_subset_rf$vital_status, SplitRatio = .7)
train_rf <- subset(data_subset_rf, sample == TRUE)
test_rf  <- subset(data_subset_rf, sample == FALSE)
train_rf$vital_status <- factor(train_rf$vital_status)
test_rf$vital_status <- factor(test_rf$vital_status)
rf <- randomForest(vital_status ~ ., data=train_rf, na.action = na.omit)
```

The following plots represent the error of the model error associated with number of trees used, as well as the importance of the different predictor variables used.

```{r, echo=FALSE}
plot(rf, main = "Random Forest")
varImpPlot(rf, main = "Importance of the Variables")
```

The following displays a confusion matrix for the predictions on the training set as well as the testing set.

```{r, echo=FALSE}
# Results:
p1 <- predict(rf, train_rf)
confusionMatrix(p1, train_rf$vital_status)
p2 <- predict(rf, test_rf)
confusionMatrix(p2, test_rf$vital_status)
```

### Neural Network

```{r, echo=FALSE}
# Neural Network:

data_subset_nn <- complete(data_imp)
#summary(data_subset_nn)
data_subset_nn <- subset(data_subset_nn, select = -c(morphology, primary_diagnosis))
#names(data_subset_nn)
```

The neural network model utilized the same predictor variables as the random forest model. However, further data processing was conducted on the dataset as dummy variables were created for each categorical feature.


```{r, echo=FALSE}
# Factored variables in 'data_subset_nn'
data_subset_nn_factored_vars <- subset(data_subset_nn,
                                       select = c(project_id,
                                                  ethnicity,
                                                  gender,
                                                  race,
                                                  vital_status,
                                                  iss_stage,
                                                  site_of_resection_or_biopsy,
                                                  tissue_or_organ_of_origin))
data_subset_nn_factored_vars_mat <- as.matrix(data_subset_nn_factored_vars)

# Non-factored variables in 'data_subset_nn'
data_subset_nn_nonfactored_vars <- subset(data_subset_nn,
                                          select = -c(project_id,
                                                      ethnicity,
                                                      gender,
                                                      race,
                                                      vital_status,
                                                      iss_stage,
                                                      site_of_resection_or_biopsy,
                                                      tissue_or_organ_of_origin))
data_subset_nn_nonfactored_vars_mat <- as.matrix(data_subset_nn_nonfactored_vars)

```


```{r, echo=FALSE}
dummies <- fastDummies::dummy_cols(data_subset_nn_factored_vars)
dummies <- subset(dummies, select = -c(1:length(names(data_subset_nn_factored_vars))))
```


```{r, echo=FALSE}
normalize <- function(matrix_x) {
  n = length(matrix_x[,1])
  return(apply(matrix_x, 2, rank) / (n + 1))
}
data_subset_nn_nonfactored_vars_mat_normalized <- normalize(data_subset_nn_nonfactored_vars_mat)
```


```{r, echo=FALSE}
data_subset_nn_normalized <- cbind(data_subset_nn_nonfactored_vars_mat_normalized, dummies)
data_subset_nn_normalized <- subset(data_subset_nn_normalized,
                                    select = -c(`vital_status_Dead`))
```


```{r, echo=FALSE}
# Renaming certain dummy-variable columns to use for the neural network model
names(data_subset_nn_normalized) <- c(
  "age_at_index",
  "days_to_birth",
  "age_at_diagnosis",
  "days_to_last_follow_up",
  "days_to_last_known_disease_status",
  "project_id_BEATAML1.0_COHORT",
  "project_id_BEATAML1.0_CRENOLANIB",
  "project_id_CGCI_BLGSP",
  "project_id_GENIE_DFCI",
  "project_id_GENIE_JHU",
  "project_id_GENIE_MDA",
  "project_id_GENIE_MSK",
  "project_id_GENIE_UHN",
  "project_id_GENIE_VICC",
  "project_id_MMRF_COMMPASS",
  "project_id_OHSU_CNL",
  "project_id_TARGET_ALL_P1",
  "project_id_TARGET_ALL_P2",
  "project_id_TARGET_ALL_P3",
  "project_id_TARGET_AML",
  "project_id_TARGET_NBL",
  "project_id_TCGA_LAML",
  "ethnicity_hispanic",
  "ethnicity_not_hispanic",
  "gender_female",
  "gender_male",
  "race_ameri_ind_or_alaska",
  "race_asian",
  "race_black_or_afr_ameri",
  "race_hawaiian_or_pacific_islander",
  "race_other",
  "race_white",
  "vital_status_Alive",
  "iss_stage_I",
  "iss_stage_II",
  "iss_stage_III",
  "site_of_resection_or_biopsy_Abdomen_NOS",
  "site_of_resection_or_biopsy_Blood",
  "site_of_resection_or_biopsy_Bone_marrow",
  "site_of_resection_or_biopsy_Bones_of_skull_and_face",
  "site_of_resection_or_biopsy_Cheek_mucosa",
  "site_of_resection_or_biopsy_Connective_and_other_soft_tissues",
  "site_of_resection_or_biopsy_Connective_and_other_soft_tissues_NOS",
  "site_of_resection_or_biopsy_Head_face_or_neck_NOS",
  "site_of_resection_or_biopsy_Hematopoietic_system_NOS",
  "site_of_resection_or_biopsy_Intra_abdominal_lymph_nodes",
  "site_of_resection_or_biopsy_Kidney_NOS",
  "site_of_resection_or_biopsy_Lip_NOS",
  "site_of_resection_or_biopsy_Liver",
  "site_of_resection_or_biopsy_Lower_gum",
  "site_of_resection_or_biopsy_Lymph_nodes_of_head_face_neck",
  "site_of_resection_or_biopsy_Mandible",
  "site_of_resection_or_biopsy_Mouth_NOS",
  "site_of_resection_or_biopsy_Ovary",
  "site_of_resection_or_biopsy_Pelvis_NOS",
  "site_of_resection_or_biopsy_Peritoneum_NOS",
  "site_of_resection_or_biopsy_Retroperitoneum",
  "site_of_resection_or_biopsy_Small_intestine_NOS",
  "site_of_resection_or_biopsy_Specified_parts_of_peritoneum",
  "site_of_resection_or_biopsy_Spleen",
  "site_of_resection_or_biopsy_Testis_NOS",
  "site_of_resection_or_biopsy_Vestibule_of_mouth",
  "tissue_or_organ_of_origin_Bone_marrow",
  "tissue_or_organ_of_origin_Bones_of_skull_and_face",
  "tissue_or_organ_of_origin_Hematopoietic_system_NOS",
  "tissue_or_organ_of_origin_Orbit_NOS"
)
```



```{r, echo=FALSE}
train_ind <- sample(seq_len(nrow(data_subset_nn_normalized)),
                    size = floor(0.7 * nrow(data_subset_nn_normalized)))

train_nn <- data_subset_nn_normalized[train_ind,]
test_nn <- data_subset_nn_normalized[-train_ind,]

# Temporarily removing numeric variables
train_nn <- subset(train_nn, select = -c(1:5))
test_nn <- subset(test_nn, select = -c(1:5))
```


```{r, echo=FALSE}
## Neural Network
## (https://datascienceplus.com/neuralnet-train-and-test-neural-networks-using-r/)
## (http://uc-r.github.io/ann_classification)

# Making the neural network model
nn <- neuralnet(vital_status_Alive ~ project_id_BEATAML1.0_COHORT + project_id_BEATAML1.0_CRENOLANIB + project_id_CGCI_BLGSP + project_id_GENIE_DFCI + project_id_GENIE_JHU + project_id_GENIE_MDA + project_id_GENIE_MSK + project_id_GENIE_UHN + project_id_GENIE_VICC + project_id_MMRF_COMMPASS +  project_id_OHSU_CNL + project_id_TARGET_ALL_P1 + project_id_TARGET_ALL_P2 + project_id_TARGET_ALL_P3 + project_id_TARGET_AML + project_id_TARGET_NBL + project_id_TCGA_LAML + ethnicity_hispanic + ethnicity_not_hispanic + gender_female + gender_male + race_ameri_ind_or_alaska + race_asian +  race_black_or_afr_ameri + race_hawaiian_or_pacific_islander + race_other + race_white + iss_stage_I + iss_stage_II + iss_stage_III + site_of_resection_or_biopsy_Abdomen_NOS + site_of_resection_or_biopsy_Blood + site_of_resection_or_biopsy_Bone_marrow + site_of_resection_or_biopsy_Bones_of_skull_and_face + site_of_resection_or_biopsy_Cheek_mucosa + site_of_resection_or_biopsy_Connective_and_other_soft_tissues + site_of_resection_or_biopsy_Connective_and_other_soft_tissues_NOS + site_of_resection_or_biopsy_Head_face_or_neck_NOS + site_of_resection_or_biopsy_Hematopoietic_system_NOS + site_of_resection_or_biopsy_Intra_abdominal_lymph_nodes + site_of_resection_or_biopsy_Kidney_NOS + site_of_resection_or_biopsy_Lip_NOS + site_of_resection_or_biopsy_Liver + site_of_resection_or_biopsy_Lower_gum + site_of_resection_or_biopsy_Lymph_nodes_of_head_face_neck + site_of_resection_or_biopsy_Mandible + site_of_resection_or_biopsy_Mouth_NOS + site_of_resection_or_biopsy_Ovary + site_of_resection_or_biopsy_Pelvis_NOS + site_of_resection_or_biopsy_Peritoneum_NOS + site_of_resection_or_biopsy_Retroperitoneum + site_of_resection_or_biopsy_Small_intestine_NOS + site_of_resection_or_biopsy_Specified_parts_of_peritoneum + site_of_resection_or_biopsy_Spleen + site_of_resection_or_biopsy_Testis_NOS + site_of_resection_or_biopsy_Vestibule_of_mouth + tissue_or_organ_of_origin_Bone_marrow + tissue_or_organ_of_origin_Bones_of_skull_and_face + tissue_or_organ_of_origin_Hematopoietic_system_NOS + tissue_or_organ_of_origin_Orbit_NOS,
                data=train_nn,
                hidden=c(50, 10),#c(50, 10),
                linear.output=FALSE,
                threshold=0.01)
```


```{r, echo=FALSE}
# Testing the resulting output
prediction_threshold <- 0.5
nn_train_results <- compute(nn, train_nn)
nn_test_results <- compute(nn, test_nn)

train_prediction <- nn_train_results$net.result
train_prediction[train_prediction >= prediction_threshold] <- 1
train_prediction[train_prediction < prediction_threshold] <- 0
train_prediction <- c(train_prediction)
train_actual <- train_nn$vital_status_Alive

test_prediction <- nn_test_results$net.result
test_prediction[test_prediction >= prediction_threshold] <- 1
test_prediction[test_prediction < prediction_threshold] <- 0
test_prediction <- c(test_prediction)
test_actual <- test_nn$vital_status_Alive

# Training set predictions
confusionMatrix(factor(train_prediction), factor(train_actual))
# Testing set predictions
confusionMatrix(factor(test_prediction), factor(test_actual))
```

# Results


# Conclusion


# Division of Work:

Yitian:

Rudra: worked on initial data preprocessing by converting empty values to `NA`, then did further preprocessing of the data for the neural network model, and finally implemented and evaluated the neural network model.

Jorge:
