---
title: "project_report"
output: 
  html_document:
    toc: true
    theme: united
date: "12/17/2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## Working directory (to be set manually)
## To ultimately remove
## setwd("C:/Users/Rudra Guin/Documents/UMich/Biostatistics_625/biostat625_final_project")
```

# Introduction/Motivation

# Preprocessing the data

Before going into analysis and modeling, we first clean up and pre-process the data

```{r}
data <- read.delim(file = 'clinical.tsv', sep = '\t', header = TRUE)
data[data == "\'--"] <- "NULL"
is.na(data) <- data == "NULL"
data <- data[, colSums(is.na(data)) < nrow(data)]
```


Viewing the unique values of the "vital status" target variable option:

```{r}
unique(data$vital_status)
```

We decided to use half of the dataset for survival analysis and the other half of the dataset for machine learning modeling and analysis. We split the dataset in half using a random selection of rows for each.

```{r}
smp_size <- floor(0.5 * nrow(data))
rand_indexes <- sample(seq_len(nrow(data)), size = smp_size)
data_subset_1 <- data[rand_indexes,] # for survival analysis
data_subset_2 <- data[-rand_indexes,] # for machine learning modeling and analysis
```

# Analysis

## Survival Analysis

We first conducted some survival analysis on the half of the initial preprocessed dataset.

```{r}

```

## Machine Learning Modeling and Analysis

We also performed some machine learning modeling and evaluated the effectiveness of the models' prediction power of the vital status of certain patients.

```{r}

```

# Conclusion
